SeedLing: Building and using a seed corpus \\ for the Human Language Project

### 2014-04-28 docs/computEL-final/concl.tex
"Universal Corpus" capitalisation to max Abney and Bird

In this paper, we have described the creation of SeedLing, a foundation text for a Universal Corpus, following the guidelines of Abney and Bird (2010; 2011). To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia. The resulting corpus is more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}. At present, SeedLing has data from 19\% of the world's living languages, covering 72\% of language families. We believe that a corpus with such diversity of languages, uniformity of format, cleanliness of data, and ease of access provides an excellent seed for a universal corpus. It is our hope that taking steps toward creating this resource will spur both further data contributions and interesting computational research with cross-linguistic or typological perspectives; we have here demonstrated SeedLing's utility for such research by using the data to perform language clustering, with promising results.

>>>>

In this paper, we have described the creation of SeedLing, a foundation text for a Universal Corpus, following the guidelines of Abney and Bird (2010; 2011). To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia. The resulting corpus is more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}. At present, SeedLing has data from 19\% of the world's living languages, covering 72\% of language families. We believe that a corpus with such diversity of languages, uniformity of format, cleanliness of data, and ease of access provides an excellent seed for a Universal Corpus. It is our hope that taking steps toward creating this resource will spur both further data contributions and interesting computational research with cross-linguistic or typological perspectives; we have here demonstrated SeedLing's utility for such research by using the data to perform language clustering, with promising results.

### 2014-04-28 docs/computEL-final/data.tex
"Universal Corpus" capitalisation to max Abney and Bird

Building on their previous paper, \newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, and we aim to adopt this structure where possible.  Two types of text are distinguished:

@@

Although data size matters in general NLP, \emph{universality} is the top priority for a universal corpus. We focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): the Online Database of Interlinear Text (ODIN), the Omniglot website, the Universal Declaration of Human Rights (UHDR), and Wikipedia.

@@

ODIN (The Online Database of Interlinear Text) is a repository of
interlinear glossed texts (IGTs) extracted from scholarly documents
\cite{lewis2006odin,lewis2010odin}.  Compared to other resources, it
is notable for the breadth of languages included and the level of
linguistic annotation.  An IGT canonically consists of three lines:
(i) the source, a sentence in a target language, (ii) the gloss, an
analysis of each source element, and (iii) the translation,
done at the sentence level. The gloss line can additionally include a
number of linguistic terms, which means that the gloss 
%can more
%properly described as being written in a metalanguage, rather than a
is written in metalanguage rather than natural language.  In ODIN, translations are into English, and glosses
are written in an English-based metalanguage.  An accepted set of
guidelines are given by the Leipzig Glossing
Rules,\footnote{\url{http://www.eva.mpg.de/lingua/resources/glossing-rules.php}}
where morphemes within words are separated by hyphens (or equal signs,
for clitics), and the same number of hyphens should appear in each
word of the source and gloss.

The data from ODIN poses the first obstacle to straightforwardly adopting Abney and Bird's data structure. 
The proposed data structure for the universal corpus is aligned at the word level, and includes a specific list of relevant features which should be used to annotate words. When we try to adapt IGTs into this format, we run into certain problems.  Firstly, there is the problem that the most fundamental unit of analysis according to the Leipzig Glossing Rules is the morpheme, not the word.  Ideally, we should encode this information explicitly in a universal corpus, assigning a unique identifier to each morpheme (instead of, or in addition to each word). Indeed, \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of \textit{word}, which undermines the central position of words in the proposed data structure.

Secondly, it is unclear how to represent the gloss.  Since the gloss line is not written in a natural language, we cannot treat it as a simple translation.  However, it is not straightforward to incorporate it into the proposed structure for analysed texts, either.  
% We could simply always map the gloss to the \textsc{gloss} field, but this would run against the suggestion that this field should "contain symbols from controlled vocabularies".  Indeed, this information might be more suited to the \textsc{morph, pos,} or \textsc{rel} fields, but it would be difficult to automatically process ODIN to determine which morphemes should be assigned to which field.
One possible resolution is to move all elements of the gloss written in capital letters to the \textsc{morph} field (as functional elements are usually annotated in this way), and all remaining elements to the \textsc{gloss} field.  However, this loses information, since we no longer know which morpheme has which meaning.  To keep all information encoded in the IGT, we need to modify \newcite{abney2011data}'s proposal.

The simplest solution we can see is to allow morphemes to be a level of structure in the universal corpus, just as documents, sentences, and words already are.  The overall architecture remains unchanged.  We must then decide how to represent the glosses.

@@

One benefit of this approach is that glossed texts are treated in exactly the same way as parallel texts.  There is a unique identifier for each morpheme, and glosses are stored under this identifier and the corresponding gloss code.  Furthermore, to motivate the important place of parallel texts in a universal corpus, Abney and Bird view translations into a high-resource reference language as a convenient surrogate of meaning. By the same reasoning, we can use glosses to provide a more detailed surrogate of meaning, only written in a metalanguage instead of a natural one.

>>>>

Building on their previous paper, \newcite{abney2011data} describe the data structure they envisage for a Universal Corpus in more detail, and we aim to adopt this structure where possible.  Two types of text are distinguished:

@@

Although data size matters in general NLP, \emph{universality} is the top priority for a Universal Corpus. We focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): the Online Database of Interlinear Text (ODIN), the Omniglot website, the Universal Declaration of Human Rights (UHDR), and Wikipedia.

@@

ODIN (The Online Database of Interlinear Text) is a repository of interlinear glossed texts (IGTs) extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.  Compared to other resources, it is notable for the breadth of languages included and the level of linguistic annotation.  An IGT canonically consists of three lines: (i) the source, a sentence in a target language, (ii) the gloss, an analysis of each source element, and (iii) the translation, done at the sentence level. The gloss line can additionally include a number of linguistic terms, which means that the gloss is written in metalanguage rather than natural language.  In ODIN, translations are into English, and glosses are written in an English-based metalanguage.  An accepted set of guidelines are given by the Leipzig Glossing Rules,\footnote{\url{http://www.eva.mpg.de/lingua/resources/glossing-rules.php}} where morphemes within words are separated by hyphens (or equal signs, for clitics), and the same number of hyphens should appear in each word of the source and gloss.

The data from ODIN poses the first obstacle to straightforwardly adopting Abney and Bird's proposal. The suggested data structure is aligned at the word level, and includes a specific list of relevant features which should be used to annotate words. When we try to adapt IGTs into this format, we run into certain problems.  Firstly, there is the problem that the most fundamental unit of analysis according to the Leipzig Glossing Rules is the morpheme, not the word.  Ideally, we should encode this information explicitly in a Universal Corpus, assigning a unique identifier to each morpheme (instead of, or in addition to each word). Indeed, \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of \textit{word}, which undermines the central position of words in the proposed data structure.

Secondly, it is unclear how to represent the gloss.  Since the gloss line is not written in a natural language, we cannot treat it as a simple translation.  However, it is not straightforward to incorporate it into the proposed structure for analysed texts, either. One possible resolution is to move all elements of the gloss written in capital letters to the \textsc{morph} field (as functional elements are usually annotated in this way), and all remaining elements to the \textsc{gloss} field.  However, this loses information, since we no longer know which morpheme has which meaning.  To keep all information encoded in the IGT, we need to modify \newcite{abney2011data}'s proposal.

The simplest solution we can see is to allow morphemes to be a level of structure in the Universal Corpus, just as documents, sentences, and words already are.  The overall architecture remains unchanged.  We must then decide how to represent the glosses.

@@

One benefit of this approach is that glossed texts are treated in exactly the same way as parallel texts.  There is a unique identifier for each morpheme, and glosses are stored under this identifier and the corresponding gloss code.  Furthermore, to motivate the important place of parallel texts in a Universal Corpus, Abney and Bird view translations into a high-resource reference language as a convenient surrogate of meaning. By the same reasoning, we can use glosses to provide a more detailed surrogate of meaning, only written in a metalanguage instead of a natural one.

### 2014-04-28 docs/computEL-final/paper.tex
Reworked discussion of experiments, trimmed down to 8 pages.
"Universal Corpus" capitalisation to max Abney and Bird

\begin{document}
\maketitle
\begin{abstract}
  A broad-coverage corpus such as the Human Language Project
  envisioned by \newcite{abney2010universal} would be a powerful
  resource for the study of endangered languages.
%, \bluenote{keep? from
 %   the perspective of both the documentary and the computational
  %  linguist}. 
  Existing corpora are limited in the range of languages covered, in
  standardisation, or in machine-readability. In this paper we present
  SeedLing, a seed corpus for the Human Language Project. We first
  survey existing efforts to compile cross-linguistic resources, then
  describe our own approach. To build the foundation text for a
  universal corpus, we crawl and clean texts from several web sources
  that contain data from a large number of languages, and convert them
  into a standardised form consistent with the guidelines of
  \newcite{abney2011data}. The resulting corpus is more
  easily-accessible and machine-readable than any of the underlying
  data sources, and, with data from 1451 languages covering 105
  language families, represents a significant base corpus for
  researchers to draw on and add to in the future. To show the utility
  of SeedLing for cross-lingual computational research, we use our
  data in the test application of detecting similar languages.
\end{abstract}

\input{intro}
\input{relwork}
%\input{liling}
\input{data}
\input{clean}
\input{cluster}
%\input{corpus}
\input{concl}

>>>>

\begin{document}
\maketitle
\begin{abstract}
  A broad-coverage corpus such as the Human Language Project
  envisioned by \newcite{abney2010universal} would be a powerful
  resource for the study of endangered languages.
%, \bluenote{keep? from
 %   the perspective of both the documentary and the computational
  %  linguist}. 
  Existing corpora are limited in the range of languages covered, in
  standardisation, or in machine-readability. In this paper we present
  SeedLing, a seed corpus for the Human Language Project. We first
  survey existing efforts to compile cross-linguistic resources, then
  describe our own approach. To build the foundation text for a
  Universal Corpus, we crawl and clean texts from several web sources
  that contain data from a large number of languages, and convert them
  into a standardised form consistent with the guidelines of
  \newcite{abney2011data}. The resulting corpus is more
  easily-accessible and machine-readable than any of the underlying
  data sources, and, with data from 1451 languages covering 105
  language families, represents a significant base corpus for
  researchers to draw on and add to in the future. To demonstrate the
  utility of SeedLing for cross-lingual computational research, we use our
  data in the test application of detecting similar languages.
\end{abstract}

\input{intro}
\input{relwork}
\input{data}
\input{clean}
\input{cluster}
\input{concl}
